---
title: The rise of populism in Europe
author: Francesco Ignazio Re
output:
  html_document:
    theme: "flatly" 
    toc: true
    toc_float: true
---
```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(maptools)
library(ggmap)
# install necessary packages
library(dplyr)
library(doBy)
# to make fortify possible without errors

library(gpclib)
gpclibPermit()
library(kableExtra)
codes <- read.csv("iso_code_europe_partial.csv", sep= ";")

codes <- codes %>% rename("Populist index" = Populist.Index )
# requires downloading and unzipping NUTS_2013_01M_SH.zip from
# http://ec.europa.eu/eurostat/web/gisco/geodata/reference-data/administrative-units-statistical-units
eurMap <- readShapeSpatial("NUTS_2013_01M_SH/data/NUTS_RG_01M_2013.shp")

# only main adm level
eurMap <- subset(eurMap, nchar(as.character(NUTS_ID)) == 2)

# EL for Grece, UK for United Kingdom
eurMapDf <- fortify(eurMap, region='NUTS_ID')

eurData <- read.csv("Overview_Elections.csv")
eurData <- eurData %>% rename("Populist index" = Populist.index ) %>% select(-X)
#eurData <- rbind(eurData,codes)
eurData <- eurData %>% mutate(`Populist index`= `Populist index` * 10)
# merge map and data
eurMapDataDf <- merge(eurMapDf, eurData, by.x="id", by.y="CountryCode")
# sort, so that polygons are drawn correctly
eurMapDataDf <- eurMapDataDf[order(eurMapDataDf$order),]

# limit data to main Europe
eurMapDataDf <- subset(eurMapDataDf, long > -15 & long < 32 & lat > 34 & lat < 75)

# add text; instead of mean I do middle (not to be to biased towards detailed coastlines)
middle = function (x) {
  (max(x) + min(x)) / 2
}  


txtVal <- summaryBy(long + lat + `Populist index` ~ id, data=eurMapDataDf, FUN=middle, keep.names=T)

# inverse order (to have visible borders)

p <- ggplot(data=eurMapDataDf) +
  geom_polygon(aes(x=long, y=lat, group=group, fill=`Populist index`)) +
  geom_path(aes(x=long, y=lat, group=group), color='black', alpha=.5) +
  geom_text(aes(x=long, y=lat, label=sprintf("%.4f", `Populist index`)), data=txtVal, col="black",cex=3,fontface =2) +
  scale_fill_gradient2(low = "navy", mid = "white", high = "red4") +
  coord_equal()+ theme(axis.line=element_blank(),axis.text.x=element_blank(),
                         axis.text.y=element_blank(),axis.ticks=element_blank(),
                         axis.title.x=element_blank(),
                         axis.title.y=element_blank())+
  geom_text(x =10 , y =59, label = "General trend of electoral outcome \n from 1992 to present")

Elections1 <- read.csv("Elections.csv", sep = ",")
Elections1 <- Elections1 %>% select(-X)

Elections00 <- Elections1 %>%filter(Year >= 1990)%>%filter(Year < 2000) %>%
              group_by(Country) %>% summarize(mean(Index)) %>%
  rename("CountryCode" = Country, "Populist index"= `mean(Index)`)
              
Elections10 <- Elections1 %>%filter(Year >= 2000)%>%filter(Year < 2010) %>%
  group_by(Country) %>% summarize(mean(Index))  %>%
  rename("CountryCode" = Country, "Populist index"= `mean(Index)`)

Elections15 <- Elections1 %>%filter(Year >= 2010)%>%filter(Year < 2015) %>%
  group_by(Country) %>% summarize(mean(Index)) %>%
  rename("CountryCode" = Country, "Populist index"= `mean(Index)`)

Elections20 <-  Elections1 %>%filter(Year >= 2015)%>%filter(Year <= 2018)%>%
  group_by(Country) %>% summarize(mean(Index)) %>%
  rename("CountryCode" = Country, "Populist index"= `mean(Index)`)


```


##Introduction

During the last decades, a new wave of right leaning populism has spread throughout European countries, fed by a rise in xenophobic and racist behaviors. These parties have dramatically grown in popularity during the last few years. They made a new turn in the way politics is perceived, as their way of communication and the topics of their agenda adjusted to what people want to hear: a simplistic interpretation of reality that exacerbates problems caused by apparently external causes such as mass immigration and cultural liberalization.

Many analysts used to believe that this rise was an effect of the economic recession of 2008-2012. However, even as Europe’s economic recovery has gathered pace, the populist surge has continued.

With that said, in this document we will try to understand first the magnitude of this phenomenon, the way inpacts the people's lives and eventually we will focus on the possible causes that generated it. 

##Analysis on the electoral outcome

For the purpose of this project, we will focus our attention on the countries that we believed witnessed this phenomenon the most, trying to catch any common pattern among them. In order to give a scale of the problem, we start by analyzing the results of the elections occured in the last twenty years. For each party that took part to every election, we assigned an index from -2 to +2, to indicate its political orientation. 

```{r tabsets, echo=FALSE} 
dt <- data.frame("FarLeft"=-2,`LeftWing`=-1,"Centre"=0,"RightWing"=1,"FarRight"=2)

kable(dt, "html") %>%
  kable_styling(bootstrap_options = "striped", font_size = 13,full_width = F) %>%
   column_spec(1:5, bold = T) %>%
  row_spec(1, bold = T, color = "white", background = "#FA725B")

```

In this way, we can get a general overview on where, politically speaking, every country leans. 

<div style="position: align; top: 0; right: 0;">

```{r , echo=FALSE,} 
p

```
</div>

On the graph above, we can observe that in the last thirty years, Germany and Uk leaned towards left-wing governments, Italy, Austria and Hungary towards right-wing governments, meanwhile France, according to the index, had a similar amount of right-wing and left-wing influences. 

## Quarterly Results {.tabset .tabset-fade .tabset-pills}

### 1990/2000
```{r , echo=FALSE} 

eurData <- Elections00

eurData <- eurData %>% mutate(`Populist index`= `Populist index` * 10)
# merge map and data
eurMapDataDf <- merge(eurMapDf, eurData, by.x="id", by.y="CountryCode")
# sort, so that polygons are drawn correctly
eurMapDataDf <- eurMapDataDf[order(eurMapDataDf$order),]

# limit data to main Europe
eurMapDataDf <- subset(eurMapDataDf, long > -15 & long < 32 & lat > 34 & lat < 75)

# add text; instead of mean I do middle (not to be to biased towards detailed coastlines)
middle = function (x) {
  (max(x) + min(x)) / 2
}  


txtVal <- summaryBy(long + lat + `Populist index` ~ id, data=eurMapDataDf, FUN=middle, keep.names=T)

# inverse order (to have visible borders)

p1 <- ggplot(data=eurMapDataDf) +
  geom_polygon(aes(x=long, y=lat, group=group, fill=`Populist index`)) +
  geom_path(aes(x=long, y=lat, group=group), color='black', alpha=.5) +
  geom_text(aes(x=long, y=lat, label=sprintf("%.4f", `Populist index`)), data=txtVal, col="black",cex=3,fontface =2) +
  scale_fill_gradient2(low = "navy", mid = "white", high = "red4") +
  coord_equal()+ theme(axis.line=element_blank(),axis.text.x=element_blank(),
                         axis.text.y=element_blank(),axis.ticks=element_blank(),
                         axis.title.x=element_blank(),
                         axis.title.y=element_blank())

p1

```




### 2000/2010

```{r,, echo=FALSE}

eurData <- Elections15

eurData <- eurData %>% mutate(`Populist index`= `Populist index` * 10)
# merge map and data
eurMapDataDf <- merge(eurMapDf, eurData, by.x="id", by.y="CountryCode")
# sort, so that polygons are drawn correctly
eurMapDataDf <- eurMapDataDf[order(eurMapDataDf$order),]

# limit data to main Europe
eurMapDataDf <- subset(eurMapDataDf, long > -15 & long < 32 & lat > 34 & lat < 75)

# add text; instead of mean I do middle (not to be to biased towards detailed coastlines)
middle = function (x) {
  (max(x) + min(x)) / 2
}  


txtVal <- summaryBy(long + lat + `Populist index` ~ id, data=eurMapDataDf, FUN=middle, keep.names=T)

# inverse order (to have visible borders)

p3 <-  ggplot(data=eurMapDataDf) +
  geom_polygon(aes(x=long, y=lat, group=group, fill=`Populist index`)) +
  geom_path(aes(x=long, y=lat, group=group), color='black', alpha=.5) +
  geom_text(aes(x=long, y=lat, label=sprintf("%.4f", `Populist index`)), data=txtVal, col="black",cex=3,fontface =2) +
  scale_fill_gradient2(low = "navy", mid = "white", high = "red4") +
  coord_equal()+ theme(axis.line=element_blank(),axis.text.x=element_blank(),
                       axis.text.y=element_blank(),axis.ticks=element_blank(),
                       axis.title.x=element_blank(),
                       axis.title.y=element_blank())

p3
```



